---
title: "Mock community bias"
author: "Rik Haagmans"
date: '2024-8-19'
output: html_document
params:
  output_dir: "../"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(params$output_dir))
library(tidyverse)
library(ggpubr)
library(ggh4x)
library(gt)
source("helper_functions.R")
source("data_files.R")
```

```{r load_data, message=FALSE}
mc_particle_quantities <- read_csv(mc_quantities_file)
mc_ref_to_virus <- read_csv(mc_ref_to_virus_file)

mc_idxstats <- load_idxstats(mc_readmap_idxstats_file, set_assembly = "mc-genomes") %>%
  left_join(
    mc_ref_to_virus,
    by = "ref"
  )

mc_idxstat_summary <- summarise_idxstats(mc_idxstats)

mc_idxstats_by_virus <- calc_expected_levels(
  sum_segment_readcounts(mc_idxstats),
  mc_particle_quantities
) %>% 
  mutate(
    base_sample = factor(base_sample, levels = base_sample_order),
    spike = factor(spike, levels = spike_order),
    rt_method = factor(rt_method, levels = method_order),
    virus = factor(virus, levels = virus_order)
  ) %>% 
  arrange(sample, rt_method, virus)

mc_expected_levels <- mc_idxstats_by_virus %>% 
        group_by(virus, rt_method) %>% 
        summarise(expected_frac = first(mc_expected_fraction))

mc_rpkm_wta_vs_sispa <- mc_idxstats_by_virus %>%
  mutate(
    rpkm = (mapped_reads / (strands * ref_length/10^3)) / (tot_reads/10^6)
    ) %>%
  pivot_wider(
    id_cols=c(base_sample, spike, virus), 
    names_from = rt_method, 
    values_from = rpkm
    )
```

## Mock community read statistics

Total and normalised number of reads mapping to MC viruses in WTA2 and SISPA
libraries:

```{r stats_mc_readmap_summary}
mc_idxstat_summary %>%
  ungroup() %>%
  select(-assembly, -sample, -percent) %>%
  mutate(total = total / 10^6) %>% 
  pivot_wider(names_from = spike, values_from = c(total, mapped, per_10k)) %>%
  arrange(base_sample, rt_method) %>%
  gt %>%
  tab_spanner(label = "Total reads (million)", starts_with("total_")) %>%
  tab_spanner(label = "Mock community reads", starts_with("mapped_")) %>%
  tab_spanner(label = "Mock community reads per 10,000 total", starts_with("per_10k_")) %>%
  fmt_number(starts_with("total_"), decimals = 2) %>%
  fmt_number(starts_with("mapped_"), decimals = 0, use_seps = TRUE) %>%
  fmt_number(per_10k_HI, decimals = 0) %>%
  fmt_number(per_10k_LO, decimals = 2) %>%
  fmt_number(per_10k_NO, decimals = 2) %>%
  cols_align(align = c("left"), columns = c(base_sample, rt_method)) %>%
  sub_missing(missing_text = "-") %>%
  cols_label(
      ends_with("NO") ~ "NO",
      ends_with("LO") ~ "LO",
      ends_with("HI") ~ "HI",
      base_sample ~ "Base sample",
      rt_method ~ "Method"
  )  
```

Fraction of reads in spiked sample WTA2 and SISPA libraries:

```{r stats_mc_readmap_summary_across_samples}
mc_idxstat_summary %>% 
  filter(base_sample %in% base_sample_faecal) %>% 
  group_by(rt_method, spike) %>% 
  summarise(
    tot_mapped = sum(mapped),
    tot_reads = sum(total)
  ) %>% 
  mutate(
    perc_mapped = round(100 * tot_mapped / tot_reads, digits = 2)
  ) %>% 
  arrange(spike, rt_method) %>% 
  gt() %>%
  fmt_percent(perc_mapped, decimals = 2, scale_values = FALSE) %>%
  cols_label(
        tot_mapped ~ "Total mapped reads",
        tot_reads ~ "Total reads",
        perc_mapped ~ "% Mapped reads",
        spike ~ "MC",
        rt_method ~ "Method"
    ) 
```

Fraction of MC reads in each sample by virus, and fold difference from expected
level:

```{r stats_mc_readmap_by_virus}
mc_idxstats_by_virus %>% 
  filter(spike != "NO") %>%
  select(base_sample, spike, rt_method, virus, mapped_reads, tot_reads, mc_expected_fraction) %>% 
  group_by(base_sample, spike, rt_method) %>% 
  mutate(perc_reads = mapped_reads / sum(mapped_reads)) %>% 
  select(-mapped_reads, -tot_reads) %>%
  pivot_wider(names_from = rt_method, values_from = perc_reads) %>% 
  # left_join(enframe(virus_host, name="virus", value="host")) %>% 
  # group_by(host) %>% 
  group_by(virus) %>%
  mutate(
    diff_WTA2 = (WTA2 - mc_expected_fraction) / mc_expected_fraction,
    diff_SISPA = (SISPA - mc_expected_fraction) / mc_expected_fraction
    ) %>% 
  gt() %>% 
  cols_label(
        base_sample ~ "Base sample",
        spike ~ "MC",
        mc_expected_fraction ~ "Expected %"
    ) %>% 
  fmt_number(starts_with("diff_")) %>% 
  fmt_percent(mc_expected_fraction, decimals = 2) %>% 
  fmt_percent(c(WTA2, SISPA), decimals = 2)
```

Number of reads mapping to reference sequences:

```{r stats_reads_mapped}
options(width = 1200)

mc_idxstats %>% 
  select(base_sample, spike, rt_method, ref, mapped_reads) %>% 
  pivot_wider(names_from = rt_method, values_from = mapped_reads) %>% 
  pivot_wider(names_from = c(base_sample, spike), values_from = c(SISPA, WTA2)) %>% 
  gt() %>% 
  tab_spanner_delim("_") %>% 
  tab_options(table.font.size = "10px", container.width = pct(100))

```

Fraction of total reads mapping to reference sequences:

```{r stats_frac_mapped}
options(width = 1200)
mc_idxstats %>% 
  select(base_sample, spike, rt_method, ref, mapped_reads, tot_reads) %>% 
  mutate(percent_mapped = mapped_reads / tot_reads ) %>% 
  select(-c(mapped_reads, tot_reads)) %>% 
  pivot_wider(names_from = rt_method, values_from = percent_mapped) %>% 
  pivot_wider(names_from = c(base_sample, spike), values_from = c(SISPA, WTA2)) %>% 
  gt() %>% 
  tab_spanner_delim("_") %>% 
  fmt_percent()%>% 
  tab_options(table.font.size = "10px", container.width = pct(100))
```

Fraction of MC reads mapping to reference sequences:

```{r stats_frac_mc_mapped}
options(width = 1200)
mc_idxstats %>% 
  select(base_sample, spike, rt_method, ref, mapped_reads, tot_mapped) %>% 
  mutate(percent_mapped = mapped_reads / tot_mapped ) %>% 
  select(-c(mapped_reads, tot_mapped)) %>% 
  pivot_wider(names_from = rt_method, values_from = percent_mapped) %>% 
  pivot_wider(names_from = c(base_sample, spike), values_from = c(SISPA, WTA2)) %>% 
  gt() %>% 
  tab_spanner_delim("_") %>% 
  fmt_percent()%>% 
  tab_options(table.font.size = "10px", container.width = pct(100))
```

Fraction of MC reads in blank sample:

```{r stats_mc_readmap_summary_blank}
mc_idxstat_summary %>% 
  filter(base_sample == "SBL") %>% 
  group_by(rt_method, spike) %>% 
  summarise(
    tot_mapped = sum(mapped),
    tot_reads = sum(total)
  ) %>% 
  mutate(
    perc_mapped = round(100 * tot_mapped / tot_reads, digits = 2)
  ) %>% 
  arrange(spike, rt_method) %>% 
  gt() %>%
  fmt_percent(perc_mapped, decimals = 2, scale_values = FALSE) %>%
  cols_label(
        tot_mapped ~ "Total mapped reads",
        tot_reads ~ "Total reads",
        perc_mapped ~ "% Mapped reads",
        spike ~ "MC",
        rt_method ~ "Method"
    ) 
```

```{r stats_mc_readmap_by_virus_blank}
mc_idxstats_by_virus %>% 
  filter(base_sample == "SBL") %>%
  select(rt_method, virus, mapped_reads, tot_reads) %>% 
  group_by(rt_method) %>% 
  mutate(perc_reads = 100 * mapped_reads / sum(mapped_reads)) %>% 
  select(-mapped_reads, -tot_reads) %>%
  pivot_wider(names_from = rt_method, values_from = perc_reads) %>% 
  left_join(enframe(virus_host, name="virus", value="host")) %>% 
  group_by(host) %>% 
  gt() %>% 
  summary_rows(
    columns = c(WTA2, SISPA), 
    fns = list(Total ~ sum(.)), 
    fmt=list(~fmt_percent(., scale_values = FALSE, decimals = 2)),
    missing_text = ""
    )%>% 
  fmt_percent(c(WTA2, SISPA), decimals = 2, scale_values = FALSE)
```


## Mock community bias 

Fraction of mock community reads of each virus, compared to expected levels:

```{r plot_mc_read_fractions}
mc_read_fraction_plot <- ggplot(
  mc_idxstats_by_virus %>% filter(spike != "NO"),
  aes(x=rt_method, y=100*(mapped_reads/tot_mapped), fill=paste0(base_sample, "-", spike))
  )+
  geom_bar(position=position_dodge(), stat="identity")+
  geom_hline(
    data=mc_expected_levels, 
    aes(yintercept = 100 * expected_frac, color = "expected")
    )+
  nice_plot()+
  facet_wrap(~virus, scales = "free_y", ncol = 4) +
  scale_y_continuous(
    limits = c(0,NA), 
    expand = expansion(add=0, mult=c(0,0.1))
    )+
  scale_color_manual(
    values = c("expected" = "blue"),
    labels = c("expected" = "Expected %")
    )+
  colorscheme_sample_fill+
  theme(
    legend.position = "bottom",
    axis.title.x = element_blank(),
    strip.text = element_text(size=9, face="bold")
    )+
  nice_legend_tight_y+
  labs(
    y = "% mock community reads",
    fill = "Sample",
    color = ""
  )
  
mc_read_fraction_plot

ggsave(
  "figures/mc_bias/mc_spiked_WTA2_vs_SISPA.png", 
  plot = mc_read_fraction_plot,
  width = 16, height = 9, dpi=300, unit="cm"
  )
```

Correlation between WTA2 and SISPA MC read fractions:

```{r plot_mc_read_fractions_wta2_vs_sispa}
ggplot(
  mc_idxstats_by_virus %>% 
    filter(spike != "NO" & base_sample != "SMC") %>% 
    mutate(frac_mc_reads = mapped_reads/tot_mapped) %>% 
    pivot_wider(id_cols = c(base_sample, spike, virus), names_from = rt_method, values_from = frac_mc_reads),
  aes(x=WTA2, y=SISPA)
  )+
  geom_smooth(method="lm", color = "black")+
  geom_point(aes(color=paste0(base_sample, "-", spike), shape = spike))+
  stat_cor(method="pearson", color = "black")+
  nice_plot()+
  facet_wrap(~virus, scales = "free", ncol = 4) +
  colorscheme_sample_color+
  theme(
    legend.position = "bottom",
    strip.text = element_text(size=9, face="bold")
    )+
  nice_legend_tight_y+
  labs(
    x = "% mock community reads WTA2",
    y = "% mock community reads SISPA",
    color = "Sample"
  )
```
Mock community virus abundance in spiked samples, comparing WTA2 to SISPA:

```{r plot_mc_abundance_spiked}
plot_limits <- c(-4, 3)

mc_spiked_wta_v_sispa_plot <- ggplot(
  mc_rpkm_wta_vs_sispa %>% filter(base_sample %in% base_sample_faecal),
  aes(x = log10(WTA2), y = log10(SISPA))
) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
    )+
  geom_point(aes(color = base_sample, shape = spike)) +
  nice_plot() +
  scale_x_continuous(limits = plot_limits) +
  scale_y_continuous(limits = plot_limits) +
  colorscheme_base_sample_color +
  scale_shape_manual(values = spike_shapes) +
  facet_wrap(~ virus, ncol = 4) +
  labs(
    x = "WTA2 abundance [log10(RPKM)]",
    y = "SISPA abundance [log10(RPKM)]",
    shape = "Spike",
    color = "Faecal sample"
  ) +
  theme(
    aspect.ratio = 1,
    legend.position = "inside",
    legend.position.inside = c(1, 0.4),
    legend.justification = c(1, 1),
    legend.box = "horizontal",
    legend.margin = margin(0,0,0,0),
    legend.box.spacing = unit(0.4, "line"),
    legend.box.margin = margin(0,0,0,0),
    strip.text = element_text(size = 9, face = "bold"),
    panel.grid = element_line(colour = "grey70", linewidth = 0.3)
  )

mc_spiked_wta_v_sispa_plot

ggsave(
  "figures/mc_bias/mc_spiked_WTA2_vs_SISPA.png",
  plot = mc_spiked_wta_v_sispa_plot,
  width = 16, height = 10, dpi=300, unit="cm"
  )
```

Mock community virus abundance in Mock Community controls, comparing WTA2 to SISPA:

```{r plot_mc_abundance_control}
mc_control_wta_v_sispa <- ggplot(
  mc_rpkm_wta_vs_sispa %>% filter(base_sample == "SMC"),
  aes(
    x = log10(WTA2),
    y = log10(SISPA),
    color = virus
  )
) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "grey23",
    linetype = "dashed"
  ) +
  geom_point(size = 1.8) +
  facet_wrap(~ spike)+
  nice_plot() +
  scale_x_continuous(limits = c(-2, 5)) +
  scale_y_continuous(limits = c(-2, 5)) +
  colorscheme_virus_color +
  theme(aspect.ratio = 1, strip.text = element_text(size = 9, face = "bold")) +
  labs(
    x = "WTA2 [log10(RPKM)]",
    y = "SISPA [log10(RPKM)]",
    color = ""
    ) +
  guides(color = guide_legend(ncol = 2, byrow = FALSE))

mc_control_wta_v_sispa

ggsave(
  "figures/mc_bias/mc_control_WTA2_vs_SISPA.png", 
  plot = mc_control_wta_v_sispa,
  width = 14, height = 5.5, dpi=300, unit="cm")
```
