---
title: "Mock Community sequencing bias"
author: "Rik Haagmans"
date: "2024-08-20"
output: html_document
params:
  output_dir: "../"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(params$output_dir))
library(tidyverse)
library(ggh4x)
library(gt)
source("helper_functions.R")
source("data_files.R")
```

```{r load_data, message=FALSE}

mc_depth <- load_depth_table(mc_readmap_depth_file)
mc_coverage <- calculate_coverage(mc_depth)

mc_depth <- left_join(
  mc_depth,
  mc_coverage %>% select(sample, ref, total_bp, covered, coverage),
  by = join_by(sample, ref)
  )  

# Calculate mean coverage graphs
mean_depth <- mc_depth %>%
  filter(coverage > 0.8) %>% # Filter out samples with low coverage
  group_by(rt_method, ref, bp) %>%
  summarise(
    depth_norm=mean(na.omit(depth_norm))
    ) %>%
  mutate(
    sample="mean",
    ref = factor(ref, levels=segment_order)
    )

mc_genome_stats <- load_sequence_stats(mc_genome_stats_file)

mc_readmap_gc <- read_tsv(
  mc_readmap_gc_file,
  col_types = cols(
    sample = col_factor(levels = full_sample_order),
    base_sample = col_factor(levels = base_sample_order),
    spike = col_factor(levels = spike_order),
    rt_method = col_factor(levels = method_order),
    replicate = col_factor(),
    ref = col_factor(levels = segment_order),
    avg_gc = col_number()
  ),
  show_col_types = FALSE
) %>%
  mutate(avg_gc = 100 * avg_gc) %>% 
  left_join(mc_genome_stats %>% select(-ref_length), by = "ref") %>%
  left_join(segment_to_virus, by = "ref") 

```

## Coverage depth

```{r mc_depth_plot_function}
plot_mc_depth_by_sample <- function(mc_depth) {
  ggplot(
    mc_depth,
    aes(
      x = bp / 1000,
      y = 10000 * depth_norm,
      color = paste0(base_sample, "-", spike)
    )
  ) +
    geom_line(alpha = .5, linewidth = 0.3) +
    nice_plot() +
    facet_wrap( ~ ref, scales = 'free', ncol = 1) +
    scale_x_continuous(expand = expansion(add = 0, mult = c(0, 0))) +
    scale_y_continuous(expand = expansion(add = 0, mult = c(0, 0))) +
    scale_color_manual(values = c(c("mean" = "#000000"), sample_palette)) +
    labs(x = "position (kbase)",
         y = expression(paste("fraction sequenced bases (", x, 10 ^ {-6}, ")")),
         color = "")
}

mc_depth_HI_samples <- mc_depth %>%
  filter(spike == "HI" &
           ref %in% c("Det7", "T5", "P22", "MHV", "M13", "RV-A_s1") &
           coverage > 0.8)
```

Coverage depth of viruses with at least 0.8X coverage in WTA2 libraries:

```{r plot_depth_by_sample_WTA2}
mc_depth_full_plot_w <-
  plot_mc_depth_by_sample(mc_depth_HI_samples %>% filter(rt_method == "WTA2"))

mc_depth_full_plot_w

ggsave(
  "figures/mc_bias/mc_depth_full_WTA2.png",
  plot = mc_depth_full_plot_w + theme(legend.position = "bottom"),
  width=16, height=20, unit="cm", dpi=300
  )
```

Coverage depth of viruses with at least 0.8X coverage in SISPA libraries:

```{r plot_depth_by_sample_SISPA}
mc_depth_full_plot_s <-
  plot_mc_depth_by_sample(mc_depth_HI_samples %>% filter(rt_method == "SISPA"))

mc_depth_full_plot_s

ggsave(
  "figures/mc_bias/mc_depth_full_SISPA.png",
  plot = mc_depth_full_plot_s + theme(legend.position = "bottom"),
  width=16, height=20, unit="cm", dpi=300
  )
```

Comparing the mean sequencing depth of WTA2 and SISPA libraries:

```{r plot_depth_mean}
mc_depth_full_sample_mean_plot <- ggplot(
  mean_depth %>% filter(ref %in% c("Det7", "T5", "P22", "M13")),
  aes(
    x = bp / 1000,
    y = 10000 * depth_norm,
    color = rt_method
  )
) +
  geom_line(
    data = mc_depth_HI_samples %>% filter(ref %in% c("Det7", "T5", "P22", "M13")),
    aes(group = base_sample),
    color = "black",
    alpha = 0.3,
    linewidth = 0.2
  ) +
  geom_line() +
  nice_plot() +
  facet_nested(ref + rt_method ~ ., scales = 'free', independent = "x") +
  scale_x_continuous(expand = expansion(add = 0, mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(add = 0, mult = c(0, 0))) +
  scale_color_manual(values = method_palette) +
  theme(
    legend.margin = margin(0, 0, 0, 0),
    strip.text = element_text(size = 9, face = "bold"),
    axis.title.x = element_text(size = 10, face = "bold")
  ) +
  labs(x = "position (kbp)",
       y = expression(paste("fraction sequenced bp (", x, 10 ^ {-6}, ")")),
       color = "")
mc_depth_full_sample_mean_plot

ggsave(
  "figures/mc_bias/mc_depth_means.png", 
  plot = mc_depth_full_sample_mean_plot + theme(legend.position = "bottom"),
  width=16, height=14, unit="cm", dpi=300
  )


```

Comparing sequencing depth uniformity:

```{r plot_depth_variation}
mc_depth_variance <- mc_depth %>%
  group_by(sample, base_sample, spike, rt_method, ref) %>%
  summarise(
    mean = mean(depth),
    sd = sd(depth),
    variance = sd ^ 2,
    cv = sd / mean,
    coverage = first(coverage)
  )

mc_dept_variance_plot <- ggplot(
  mc_depth_variance %>%
    filter(spike == "HI"),
  aes(
    y = ref,
    x = cv,
    color = rt_method,
    shape = base_sample
  )
) +
  geom_point() +
  nice_plot() +
  scale_y_discrete(limits = rev(segment_order))+
  scale_color_manual(values = method_palette) +
  scale_shape_manual(values = base_sample_shapes) +
  xlim(0, 2.4) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.margin = margin(0, 0, 0, 0),
    axis.title.y = element_blank()
  ) +
  labs(x = "coefficient of variation", color = "", shape = "")

mc_dept_variance_plot

ggsave(
  "figures/mc_bias/mc_depth_variance.png",
  plot = mc_dept_variance_plot +
    guides(shape = guide_legend(nrow = 2)) +
    guides(color = guide_legend(nrow = 2)),
  width = 7,
  height = 10,
  unit = "cm",
  dpi = 300
)
```

## GC Content

Variation in average GC content in relation to genome coverage:

```{r plot_gc_deviation_vs_coverage}
mc_gc_vs_cov_plot <- ggplot(
  left_join(
    mc_readmap_gc %>% filter(spike %in% c("HI", "LO")),
    mc_coverage %>% select(sample, ref, coverage),
    by = join_by(sample, ref)
  ),
  aes(
    x = coverage,
    y = avg_gc - ref_gc,
    color = ref,
    shape = base_sample
  )
) +
  geom_vline(xintercept = .2,
             color = "grey23",
             linetype = "dashed") +
  geom_point() +
  nice_plot() +
  colorscheme_virus_color +
  scale_shape_manual(values = base_sample_shapes) +
  facet_wrap( ~ rt_method) +
  labs(color = "virus", y = "percentage points diference\nfrom genome %GC")

mc_gc_vs_cov_plot

ggsave(
  "figures/mc_bias/mc_gc_vs_cov.png",
  plot = mc_gc_vs_cov_plot,
  width = 16, height = 8, units = "cm", dpi = 300
)

```

Deviation of average GC content compared to actual genomic GC content:

```{r plot_gc_deviation}
ref_labels <- mc_genome_stats %>%
  select(-ref_length) %>%
  mutate(ref_label = paste0(ref, "\n(", ref_gc, "% GC)")) %>%
  pull(ref_label, name = ref)

mc_readmap_gc_highcov <- left_join(
  mc_readmap_gc,
  mc_coverage %>% select(sample, ref, coverage),
  by=join_by(sample, ref)
  ) %>%
  filter(coverage > 0.2)

mc_gc_bias_plot <- ggplot(
  mc_readmap_gc_highcov %>%
    filter(virus %in% c("T5", "Det7", "P22", "M13", "MHV", "BVDV")), 
  aes(x = rt_method, y = avg_gc - ref_gc)
  ) +
  geom_hline(yintercept = 0,
             color = "grey23",
             linetype = 'dashed') +
  geom_point(aes(color = virus, shape = base_sample)) +
  facet_wrap(
    ~ ref,
    scales = "free_y",
    ncol = 3,
    labeller = as_labeller(ref_labels)
  ) +
  nice_plot() +
  scale_y_continuous(limits = c(-2.5, 9)) +
  scale_color_manual(values = virus_palette, guide = "none") +
  scale_shape_manual(values = base_sample_shapes) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 0),
    legend.justification = c(1, 0),
    legend.direction = "vertical",
    strip.text = element_text(size = 9, face = "bold")
  ) +
  labs(x = "",
       y = "percentage points difference\nfrom genome GC",
       shape = "")

mc_gc_bias_plot

mc_gc_bias_RV_plot <- ggplot(
  mc_readmap_gc_highcov %>% filter(virus == "RV-A"),
  aes(x = ref, y = avg_gc)
  ) +
  geom_point(aes(shape = base_sample)) +
  geom_point(aes(y = ref_gc, x = ref),
             color = "red",
             shape = 4) +
  facet_grid( ~ rt_method, scales = "free", space = "free_x") +
  nice_plot() +
  scale_shape_manual(values = base_sample_shapes) +
  theme(strip.text = element_text(size = 9, face = "bold")) +
  rotate_x_text()+
  labs(x = "",
       y = "% GC",
       shape = "")

mc_gc_bias_RV_plot

ggsave(
  "figures/mc_bias/mc_gc_bias_non-RV-A.png",
  plot = mc_gc_bias_plot,
  width = 9, height = 10, units = "cm", dpi = 300
)

ggsave(
  "figures/mc_bias/mc_gc_bias_RV-A.png",
  plot = mc_gc_bias_RV_plot,
  width = 16, height = 8, units = "cm", dpi = 300
)
```

```{r stats_non_rva}
mc_readmap_gc_highcov %>% 
  filter(virus %in% c("T5", "Det7", "P22", "M13", "MHV", "BVDV")) %>% 
  select(sample, ref, avg_gc, ref_gc) %>% 
  pivot_wider(id_cols = c(ref, ref_gc), names_from = sample, values_from = avg_gc) %>% 
  arrange(ref_gc) %>% 
  gt() %>% 
  fmt_percent(-ref, decimals = 2, scale_values = FALSE) %>% 
  tab_spanner(label = "WTA2", ends_with("-WTA2")) %>% 
  tab_spanner(label = "SISPA", ends_with("-SISPA"))


```



```{r stats_rva}
mc_readmap_gc_highcov %>% filter(virus == "RV-A") %>% 
  select(sample, ref, avg_gc, ref_gc) %>% 
  pivot_wider(id_cols = c(ref, ref_gc), names_from = sample, values_from = avg_gc) %>% 
  arrange(ref_gc) %>% 
  gt() %>% 
  fmt_percent(-ref, decimals = 2, scale_values = FALSE) %>% 
  tab_spanner(label = "WTA2", ends_with("-WTA2")) %>% 
  tab_spanner(label = "SISPA", ends_with("-SISPA"))
  

```